---
title: "Bayesian Final Project"
format: html
---

# Motivation/Goals

Better understand how rule changes in baseball have affected gameplay.

Links:
- https://www.mlb.com/news/history-of-mlb-rules-changes

Changes:
- **designated hitter**
   - AL: 1973, home team: 1986, universal: 2022
- 3 man pitcher rule
- longer base distances

Throughout the history of baseball, different rule changes have been made for 

The motivation of this project is to determine how the introduction of designated hitters in baseball affected batting statistics.

This project surrounds causal analysis, which is included in Chapter 6 of the textbook.

## Running Code

The data used in this project come from Sean Lahman's Baseball Database.

Lahman, S. (2023) Lahman's Baseball Database, 1871-2022, Main page, https://www.seanlahman.com/baseball-archive/statistics/. This dataset includes baseball statistics in Major Leauge Baseball for the seasons ranging from 1871-2022.

```{r setup}
pacman::p_load(baseballDBR, Lahman, rethinking, tidyverse)
set_cmdstan_path("C:\\Users\\radma\\Downloads\\cmdstan-2.33.1")
data("Batting")
data("Fielding")
```

There will be two different sets: an old introduction in the American League while the National League still banned designated hitters, and a new introduction when the MLB introduced universal designated hitting.

Now, we will create two partitions of data - both batting statistics, one of designated hitters and the other of other hitters.

```{r}
DH_List <- Appearances %>%
  filter(G_dh/G_all >= 0.5, G_all >= 30, 
         yearID >= 1956, yearID < 1981)
DH_Batting <- merge(Batting, DH_List, by = c("playerID", "yearID")) %>%
  drop_na()
DH_Batting$SLG <- SLG(DH_Batting)
DH_Batting$OBP <- OBP(DH_Batting)
DH_Batting$DH <- 2

no_DH_List <- Appearances %>%
  filter(G_dh/G_all == 0, G_all >= 30, 
         yearID >= 1956, yearID < 1981)
no_DH_Batting <- merge(Batting, no_DH_List, by = c("playerID", "yearID")) %>%
  drop_na()
no_DH_Batting$SLG <- SLG(no_DH_Batting)
no_DH_Batting$OBP <- OBP(no_DH_Batting)
no_DH_Batting$DH <- 1

batstats <- rbind(DH_Batting, no_DH_Batting) %>%
  drop_na()

# rm(DH_List, no_DH_List)
```

```{r}
ggplot() + 
  geom_density(aes(DH_Batting$OBP), color = "red") + 
  geom_density(aes(no_DH_Batting$OBP), color = "blue")
```

```{r}
table(batstats$DH)
print("Non-DH:")
fivenum(batstats$OBP[batstats$DH == 1])
sd(batstats$OBP[batstats$DH == 1])
print("DH:")
fivenum(batstats$OBP[batstats$DH == 2])
sd(batstats$OBP[batstats$DH == 2])
```

```{r}

# adding dummy variables: pre-post DH era, 
batstats$DHera <- rep(1, nrow(batstats))

for (i in 1:nrow(batstats))
{
  if (batstats$yearID[i] >= 1973)
    batstats$DHera[i] = 2
}

```

```{r}
ggplot() + 
  geom_density(aes(batstats$OBP[batstats$DHera == 1]), color = "blue") + 
  geom_density(aes(batstats$OBP[batstats$DHera == 2]), color = "red")

ggplot() + 
  geom_density(aes(batstats$HR[batstats$DH == 1]/batstats$G[batstats$DH == 1]), color = "blue") + 
  geom_density(aes(batstats$HR[batstats$DH == 2]/batstats$G[batstats$DH == 2]), color = "red")
```



```{r}
table(batstats$DH)
print("Non-DH:")
fivenum(batstats$HR[batstats$DH == 1]/batstats$G[batstats$DH == 1])
sd(batstats$HR[batstats$DH == 1])
print("DH:")
fivenum(batstats$HR[batstats$DH == 2]/batstats$G[batstats$DH == 2])
sd(batstats$HR[batstats$DH == 2])
```


# Regression

We will be using linear regression.
```{r}
precis(batstats)
```

```{r}
per_AB <- function(x)
{
  list <- c()
  for (i in 1:length(x))
  {list <- c(list, x[i]/batstats$AB[i])}
  return(list)
}


batstats$H_perAB <- per_AB(batstats$H)
batstats$R_perAB <- per_AB(batstats$R)
batstats$X2B_perAB <- per_AB(batstats$X2B)
batstats$X3B_perAB <- per_AB(batstats$X3B)
batstats$HR_perAB <- per_AB(batstats$HR)
batstats$RBI_perAB <- per_AB(batstats$RBI)
batstats$BB_perAB <- per_AB(batstats$BB)
batstats$SO_perAB <- per_AB(batstats$SO)
batstats$IBB_perAB <- per_AB(batstats$IBB)


dat <- batstats %>%
  select(DH,
         yearID,
         DHera,
         lgID.x,
         stint,
         G,
         AB,
         SLG,
         OBP,
         H_perAB,
         R_perAB,
         X2B_perAB,
         X3B_perAB,
         HR_perAB,
         RBI_perAB,
         BB_perAB,
         SO_perAB,
         IBB_perAB)

dat <- dat %>% rename(lgID = lgID.x)

# rm(old_AL, old_NL, Fielding, battingLabels, Batting, DH_Batting, no_DH_Batting, cat, vars, i, list, c, batstatsH, per_game)

```

# What to search to get to right chapter:
Interpreting these parameters is easy enough—they are the expected heights in each category

```{r}
precis(dat)
```

```{r}
hist(dat$HR_perAB)
ggplot() + 
  geom_density(aes(dat$HR_perAB[dat$DH == 1]), color = "blue") + 
  geom_density(aes(dat$HR_perAB[dat$DH == 2]), color = "red") + 
  geom_density(aes(dat$HR_perAB), color = "black")
```

```{r}
hist(dat$OBP)
```


```{r}
dat$DH_dummy <- dat$DH - 1
dat$OBP_std <- standardize(dat$OBP)

obp_lin_mod <- quap(
  alist(
    OBP_std ~ dnorm(mu, sigma),
    mu <- a + bDH*DH_dummy,
    a ~ dnorm(-0.05, 0.5),
    bDH ~ dnorm(0.5, 0.5),
    sigma ~ dexp(1)), 
  data = dat)

labels <- paste("a[" , 1:2 , "]:", levels(dat$DH), sep = "")
plot(precis(obp_lin_mod, depth = 2 , pars = "a"), labels = labels,
     xlab = "expected OBP (std)")

precis(obp_lin_mod, depth = 2)

post <- extract.samples(obp_lin_mod)
precis(post, depth = 2)
```

```{r}
xseq <- seq( from=min(dat$DH_dummy)-0.15 , to=max(dat$DH_dummy+0.15 , length.out=30 ))
mu <- link(obp_lin_mod, data=list(DH_dummy=xseq))
mu_mean <- apply(mu,2,mean)
mu_PI <- apply(mu,2,PI)


plot( OBP_std ~ DH_dummy , data=dat )
lines( xseq , mu_mean , lwd=2 )
shade( mu_PI , xseq )
```


```{r}
obp_quap <- quap(
  alist(
    OBP_std ~ dnorm(mu, sigma),
    mu <- a[DH],
    a[DH] ~ dnorm(0, 2),
    sigma ~ dexp(1)), 
  data = dat)

labels <- paste("a[" , 1:2 , "]:", levels(dat$DH), sep = "")
plot(precis(obp_quap, depth = 2 , pars = "a"), labels = labels,
     xlab = "expected OBP (std)")

precis(obp_quap, depth = 2)
```



<!-- ```{r} -->
<!-- dat$OBP_std <- standardize(dat$OBP) -->
<!-- obp_mod <- quap( -->
<!--   alist( -->
<!--     OBP_std ~ dnorm(mu, sigma), -->
<!--     mu <- a[DH], -->
<!--     a[DH] ~ dnorm(1, 2), -->
<!--     sigma ~ dexp(1)),  -->
<!--   data = subset(dat, DHera == 2))  -->

<!-- labels <- paste("a[" , 1:2 , "]:", levels(dat$DH), sep = "") -->
<!-- plot(precis(obp_mod, depth = 2 , pars = "a"), labels = labels, -->
<!--      xlab = "expected OBP (std)") -->

<!-- precis(obp_mod, depth = 2) -->

<!-- post <- extract.samples(obp_mod) -->
<!-- post$diff_fm <- post$a[,2] - post$a[,1] -->
<!-- precis(post, depth = 2) -->
<!-- ``` -->

```{r}
min(dat$OBP_std)
max(dat$OBP_std)
```


```{r}
post <- extract.samples(obp_mod)
post$diff_fm <- post$a[,2] - post$a[,1]
precis(post, depth = 2)
```


```{r}
mod_ulam <- ulam(alist(
OBP_std ~ dnorm(mu, sigma),
    mu <- a + bDH*DH_dummy,
    a ~ dnorm(-0.05, 0.5),
    bDH ~ dnorm(0.5, 0.5),
    sigma ~ dexp(1)), 
  data = dat, chains = 4,
start = list(a = 0.2, bdh = 0.3, sigma = 0.1))
```

```{r}
precis(mod_ulam, depth = 2)
plot(mod_ulam, depth = 2)
```


# Additional Code
```{r}
noDH_AL_RPY <- noDH_AL_Batting %>% 
  filter(lgID == "AL") %>%
  group_by(yearID) %>%
  summarise(R = sum(R))

noDH_NL_RPY <- noDH_NL_Batting %>% 
  filter(lgID == "NL") %>%
  group_by(yearID) %>%
  summarise(R = sum(R))

DH_AL_RPY <- DH_AL_Batting %>% 
  filter(lgID == "AL") %>%
  group_by(yearID) %>%
  summarise(R = sum(R))

DH_NL_RPY <- DH_NL_Batting %>% 
  filter(lgID == "NL") %>%
  group_by(yearID) %>%
  summarise(R = sum(R))

```

```{r}
old_AL_preDH <- Batting %>%
  filter(lgID == "AL", yearID >= 1968, yearID < 1973) %>%
  drop_na()
old_AL_preDH$prepost <- rep(0, nrow(old_AL_preDH))
old_AL_postDH <- Batting %>%
  filter(lgID == "AL", yearID >= 1973, yearID < 1978) %>%
  drop_na()
old_AL_postDH$prepost <- rep(1, nrow(old_AL_postDH))
old_AL <- rbind(old_AL_preDH, old_AL_postDH)
rm(old_AL_preDH, old_AL_postDH)

old_NL_preDH <- Batting %>%
  filter(lgID == "NL", yearID >= 1968, yearID < 1973) %>%
  drop_na()
old_NL_preDH$prepost <- rep(0, nrow(old_NL_preDH))
old_NL_postDH <- Batting %>%
  filter(lgID == "NL", yearID >= 1973, yearID < 1978) %>%
  drop_na()
old_NL_postDH$prepost <- rep(1, nrow(old_NL_postDH))
old_NL <- rbind(old_NL_preDH, old_NL_postDH)
rm(old_NL_preDH, old_NL_postDH)
```

